A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-12-01 bioc-workflow.Rnw
update author contribs
Merging lisa's changes

In this workflow, we will describe how to prepare the spatial
proteomics data starting from a spreadsheet containing quantitative
mass spectrometry data. We will focus on a recent pluripotent mouse
embryonic stem cells experiment \cite{hyper}. These data, as well as
additional annotated and pre-formatted datasets from various species
are readily available in the \Biocexptpkg{pRolocdata} package.

@@

@@

To make use of the full functionality of the \Biocpkg{pRoloc} software
one needs to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for
the efficient manipulation and processing of mass spectrometry and
proteomics data in R. Figure 1 illustrates a simplified view of the
\texttt{MSnSet} structure; there exists 3 key sub-parts (termed slots)
to such a data object: (1) the \texttt{exprs} (short for
\textit{expression} data) slot for storing the quantitation data, (2)
the \texttt{fData} slot (short for \textit{feature}-metadata) for
storing the feature meta-data, and finally (3) the \texttt{pData} slot
(short for \textit{pheno}-metadata, i.e. sample phenotypic data) for
storing the sample meta-data.

@@

\subsection*{Adding user-defined marker lists}
It is also possible for users to use their own marker list with the
\texttt{addMarkers} function. The user needs to create a named vector
of marker localisation, or a create a csv file with two columns (one
for the protein names, one for the corresponding sub-cellular marker
annotation) and pass the vector or filename respectively to the
function. As previously mentioned, the protein names of these markers
must match some (but not necessarily all) of the \texttt{MSnSet}'s
feature names. See \texttt{?addMarkers} for more details.

@@

\subsection*{Visualising markers}
Having added a the mouse markers to our \texttt{fData} from the \texttt{pRolocmarkers} we can now visualise these annotations along the PCA plot using the
\texttt{plot2D} function and then use the \texttt{addLegend} function
to map the marker classes to the pre-defined colours. We also display
the data along the first and seventh PCs using the \texttt{dims}
argument. Note that in these calls to the \texttt{plot2D} function, we
have omitted the \texttt{fcol} argument and use of the default
\texttt{"markers"} feature variable to annotate the plot. We choose
to display PCs 1 and 7 to illustrate that while upper principal
components explain much less variability in the data (2.23\% for PC7,
as opposed to 48.41\% for PC1), we see that the mitochondrial (purple)
and peroxisome (dark blue) clusters can be differenciated, despite the
apparent overlap in the visualisation of the two first PCs
(Figure \ref{fig:plotmarkers}).

@@

\subsection*{Features of interest}

In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object. For example, if we wanted to highlight a proteins with the
accession numbers Q8CG48, Q8CG47, Q8K2Z4, and Q8C156, which are some
of the proteins known to form part of the 13S condensin complex, we
would create a first create a \texttt{FeaturesOfInterest} object, and
subsequently highlight their location on a PCA plot with the
\texttt{highlightOnPlot} function.

@@

plot2D(hl)
addLegend(hl, cex = .6)
highlightOnPlot(hl, foi13s)
@  
  \caption{Highlighting protein features of interest.}
  \label{fig:foi}
\end{figure}

>>>>

In this workflow, we will describe how to prepare the spatial
proteomics data starting from a spreadsheet containing quantitative
mass spectrometry data, through to some essential data processing
steps, and finish with different applications of machine learning
(Figure \ref{fig:overview}). We focus on a recent pluripotent mouse
embryonic stem cells experiment \cite{hyper}. These data, as well as
additional annotated and pre-formatted datasets from various species
are readily available in the \Biocexptpkg{pRolocdata} package.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/overview.png}
  \caption{Schematic overview of the pRoloc pipeline from data import, through to data processing, machine learning and data export.}
  \label{fig:overview}
\end{figure}

@@

To make use of the full functionality of the \Biocpkg{pRoloc} software
one needs to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for
the efficient manipulation and processing of mass spectrometry and
proteomics data in R. Figure \ref{fig:msnset} illustrates a simplified view of the
\texttt{MSnSet} structure; there exists 3 key sub-parts (termed slots)
to such a data object: (1) the \texttt{exprs} (short for
\textit{expression} data) slot for storing the quantitation data, (2)
the \texttt{fData} slot (short for \textit{feature}-metadata) for
storing the feature meta-data, and finally (3) the \texttt{pData} slot
(short for \textit{pheno}-metadata, i.e. sample phenotypic data) for
storing the sample meta-data.

@@

\subsection{*Adding user-defined marker lists}
It is also possible for users to use their own marker list with the
\texttt{addMarkers} function. The user needs to create a named vector
of marker localisation, or a create a csv file with two columns (one
for the protein names, one for the corresponding sub-cellular marker
annotation) and pass the vector or filename respectively to the
function. As previously mentioned, the protein names of these markers
must match some (but not necessarily all) of the \texttt{MSnSet}'s
feature names. See \texttt{?addMarkers} for more details.

@@

\subsection{*Visualising markers}
Having added a the mouse markers to our \texttt{fData} from the
\texttt{pRolocmarkers} we can now visualise these annotations along
the PCA plot using the \texttt{plot2D} function and then use the
\texttt{addLegend} function to map the marker classes to the
pre-defined colours. We also display the data along the first and
seventh PCs using the \texttt{dims} argument. Note that in these calls
to the \texttt{plot2D} function, we have omitted the \texttt{fcol}
argument and use of the default \texttt{"markers"} feature variable to
annotate the plot. We choose to display PCs 1 and 7 to illustrate that
while upper principal components explain much less variability in the
data (2.23\% for PC7, as opposed to 48.41\% for PC1), we see that the
mitochondrial (purple) and peroxisome (dark blue) clusters can be
differenciated, despite the apparent overlap in the visualisation of
the two first PCs (Figure \ref{fig:plotmarkers}).

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have a specific subset of proteins
they wish to highlight in their data they would first create a
\texttt{FeaturesOfInterest} object. For example, if we wanted to
highlight a proteins with the accession numbers Q8CG48, Q8CG47,
Q8K2Z4, and Q8C156, which are some of the proteins known to form part
of the 13S condensin complex, we would create a first create a
\texttt{FeaturesOfInterest} object then use the function
\texttt{highlightOnPlot} to highlight their location on a PCA plot.

@@

plot2D(hl)
addLegend(hl, cex = .6)
highlightOnPlot(hl, foi13s)
@  
  \caption{Highlighting protein features of interest}
  \label{fig:foi}
\end{figure}
